---
title: "Analysis of changes in the price of salmon and how it affects the share prices of norwegian salmon producing companies "
author: "Even Grindstad Brovold & Orjan Mikal Antonsen"
date: "15 november 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(httr)
library(rjstat)
library(jsonlite)
library(PxWebApiData)
library(tidyr)
library(readr)
library(dplyr)
library(ggvis)
library(lubridate)
library(mosaic)
```

#Introduction
The norwegian salmon has become a highly demanded commodity, during the past couple of decades. This had made salmon production a big buisiness in Norway. In fact the fifth most valuable company listed on the norwegian stock marked, is a salmon producing company, respectively "Marine harvest". The second largest salmon producing company in Norway is "Salmar", who holds the tenth place on the list of most valuable stocks. Considering a rich country like Norway, the position of these two companies means they are doing serious buisiness. 

In this analysis we are going to take a look at how the stock prices of these two have changed the past years. We will also take a look at how the price of salmon has changed during the course of the last years. Then we will take a look on how the changes in the price of salmon, has affected the stock value of the aforementioned salmon producing companies. We will be using data from the statistical bureau in Norway "SSB", and from https://www.netfonds.no/ who provides stock data. As addidtional information, we will be focusing on raw salmon and not frozen. This is simply due to the fact that there is fourthy times more turnover of raw salmon than frozen.


##The salmon price during the last years

```{r}
laksepris <- ApiData("http://data.ssb.no/api/v0/dataset/1122.json?lang=no", 
                     getDataByGET = TRUE, col_types = cols(måned = col_date(format = "%Y%m")))

# Henter ut datasettet fra laksepris
laksepris <- laksepris[[1]]


# Long til wide
laksepris <- spread(laksepris, statistikkvariabel, value)

colnames(laksepris) <- c("Goods", "Date", "Price_per_kg_NOK", "Volume_tons")

laksepris <- laksepris %>% 
  filter(Goods == "Fersk oppalen laks")

#Fikser datoen
today <- Sys.Date()

newdate <- seq(as.Date("2000-01-01"), as.Date(today), by = "weeks")

# Oppdateringen fra SSB henger en uke etter slik at vi må fjerne den siste uken fra,
# newdate før vi klistrer den inn i datokolonnen i laksepris.

newdate <- head(newdate, -1)

laksepris$Date <- newdate

# I tilfeller hvor dato matcher med laksepris datoene, betyr det at oppdatering
# fra SSB i nåværende øyeblikk ikke henger en uke etter, slik at da ser man bort ifra
# de to siste linjenene, og kjører rm(list=ls()) som står øverst i r fila.
# Deretter kjører man laksepris$Datw <- newdate alene, for å få riktig dato.


#Datoene strekker seg ikke like langt tilbake i tid,
# som salmar_marine så tar med relevante datoer.

laksepris <- filter(laksepris, Date > "2007-05-04")
fixeddate <- seq(as.Date("2007-05-07"), as.Date(today), by = "weeks")

fixeddate <- head(fixeddate, -1)

laksepris$Date <- fixeddate

rm(newdate)

laksepris %>% ggvis(~Date, ~Price_per_kg_NOK) %>%
  layer_paths() %>%
  add_axis("y", title = "Price per kg in NOK") %>%
  layer_smooths(stroke := "red")

```

We will start this analysis by diving into the develpment of the salmon price during the past years. As we can see from the graph above, the price has almost tripled since 2007 to 2018. Then later the prize has taken a litle hit at start of the summer of 2018. There has been massive fluctuations in the price, especially in the years 2011-2012, 2014-2015 and 2017-2018. The reason behind these fluctuations is not what this analysis will focus on, so this is merely an observation. The most important lesson from this graph is the red line showing a continuosly growth in the salmon price in the past years.
